Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-07-22 review_article.Rmd

Utilising maps to present statistics has been widely used for centuries.
These maps connect data and/or statistics to the geographical representation of areas that are generally familiar to the audience. However, it is not enough to be recognisable, if the meaning of the statistic cannot be seen. This is particularly relevant to cancer statistics, as cancer outcomes directly relate to the people living within a geographical area. In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer statistics. This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes, and health outcomes more broadly.

@@

@@

<!--
Disease maps can rapidly highlight geographical
patterns that would otherwise be hidden in a
table, and are commonly used for hypothesis
generating and as the first step in exploratory
spatial data analysis. @SE
Geography is concerned with the identification and
explanation of spatial structure, pattern, and process,
and with the analysis and explanation of the links
between humans and the environment (2). @SAMGIS

@@

<!-- what is a choropleth map -->
A choropleth map is used to show the spatial relationship, or variability, of measurements. Created as a storage device which preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO]. They are constructed by drawing the given geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI]. Early versions used symbols or patterns instead of colour. @CPISACA discusses the use of choropleth maps for visualising cancer data, and @DMAHP gives an overview of the development of the use of choropleth maps for displaying disease data. A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form. Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

@@

<!--
Certain limitations to the standard graphic techniques have also become apparent, however.
They are slow, and do not permit instant retrieval of information. They are untrustworthy when interpretations of spatial association are needed; there are both severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually.

These opportunities will be lost if locational coding is not inserted into the data referencing system as such banks are built, however.

of some element, for example population density, or purchasing power, or agricultural fertility; reality, of course, is not like this. the background variable assumed to be uniform (such as population density) can be adjusted so that it behaves as if uniformity were the case
-->

@@

<!--
Literature regarding measures presented in public facing atlases.

@@

@@

A grey literature review conducted by (ref) identified 33 publicly available cancer atlases published on the internet between 01/01/2010 to 11/11/2015. Reflecting the high use of chloropleth maps founds in the literature, outlined in the section above, all the public facing maps identified in this review were also chloropleth maps.

### Publishers {-}

The majority of atlases were published by non-commercial organisations, including not-for-profits (NFPs), government, research organisations, advocacy groups or a partnership between an NFP & government. Only one map was published by a commercial entity (Maps of Cancer Mortality Rates in Spain), in this case a media organisation El Pais^[https://elpais.com/].

### Reported measures {-}

The majority of maps identified in this study reported age adjusted rates of either incidence, mortality or survival. The list below summarises these measures, and provides a definition of each measure.

### Geographical coverage {-}

Identified cancer atlases covered geographies from all around the world: 4 were global, 3 from Australia (AUS), 11 from the United States (US), 2 from Canada (CAN), 7 from the United Kingdom (UK), 2 from Spain, 1 from Switzerland, 1 from Germany, 1 from Norway, and 1 covering the European Union. Not all maps had a national focus and 10 covered a region or state rather than an entire nation. The states or counties/regions covered were South Australia (AUS), Queensland (AUS), Ontario (CAN), Valencia (Spain), Pennsylvania county Massachusetts (US), New Hampshire (US), Cape Cod (US), Missouri (US), Florida (US), New York State (US) and Arizona (US).

\newpage
Table: (\#tab:measures) Measures used to report cancer statistics

### Statistical uncertainty

Cancer atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map. Maps that only reported this information within the supplementary material were not considered to have directly attempted to report uncertainty.

>>>>

@@

Maps have been adopted to present geospatial statistics for centuries.
These maps connect data and/or statistics to the geographical representation of areas that are generally familiar to the audience.
However, it is not enough for areas on maps to be recognisable, if the distribution of the statistic cannot be understood.
This is particularly relevant to cancer statistics, as cancer outcomes directly relate to the people living within a geographical area.
In situations where the people are of interest, not the land they live on, it is reasonable to explore views that enhance the communication of the cancer statistics.
This has spurred innovations over the previous centuries to enhance the shapes and maps presented to effectively communicate cancer outcomes, and health outcomes more broadly.

@@

<!--
Disease maps present geographic patterns that may have been overlooked in a table that obscures the geospatially related statistics [@SAMGIS].
Often used as a first step to drive hypotheses

Transforming statistics to visualisations begins with the coordinates of observations.
The observation points can be aggregated to regions within boundaries.
The polygon regions may be presented directly, or transformed first according to associated values.

The visualisation method used will depend on the intended use of the map, namely the message a map will communicate.

commonly used for hypothesis generating and as the first step in exploratory
spatial data analysis. @SE Geography is concerned with the identification and
explanation of spatial structure, pattern, and process, and with the analysis and explanation of the links
between humans and the environment (2). @SAMGIS

@@

<!-- what is a choropleth map -->
A choropleth map is used to show the spatial characteritics of a 
relationship, or variability, of measurements. Created as a storage device which preserves locations for geographically ordered data, with usage dating back to the 1800s [@GOINO]. They are constructed by drawing the given geographic or political boundaries, and filling the shapes with colours to represent values of a measured variable [@EI]. Early versions used symbols or patterns instead of colour. @CPISACA discusses the use of choropleth maps for visualising cancer data, and @DMAHP gives an overview of the development of the use of choropleth maps for displaying disease data. A choropleth map is a true map of the topology, constructed for visual inspection of spatial patterns across a familiar geographic form. Figure 1 shows a choropleth map where each state of the United States of America is coloured by the average annual rate of new cases of lung and bronchus cancer from years 2012 to 2016.

@@

<!--
criticisms:
Certain limitations to the standard graphic techniques have also become apparent, however.
They are slow, and do not permit instant retrieval of information. They are untrustworthy when interpretations of spatial association are needed; there are both severe physiological and psychological constraints on our abilities to assess correlations of spatial distributions visually.
-->

@@

<!--
Literature regarding measures presented in public facing atlases.
-->

### Reported measures in cancer atlases

The list below summarises these measures, and provides a definition of each measure.

Table: (\#tab:measures) Measures used to report cancer statistics

@@

### Supporting material in cancer atlases

<!-- Moved Jessies stat uncertainty up-->
Cancer atlases were considered to report uncertainty to the non-expert user if they included a measure of statistical uncertainty either within or alongside the map. Maps that only reported this information within the supplementary material were not considered to have directly attempted to report uncertainty.

@@

Cancer maps are effective communication tools for a general or non-expert audience and are commonly used in the public domain to communicate the output of sophisticated statistical analyses to these audiences. The heavy use of chloropleth maps within the research literature is reflected in the types of maps that are found in the public domain. A grey literature review conducted by (ref) identified 33 cancer atlases published on the internet between 01/01/2010 to 11/11/2015 all of which were chloropleth maps. These maps were mostly published by non-commercial organisations, including not-for-profits (NFPs), government, research organisations, advocacy groups or a partnership between an NFP & government. Only one map was published by a commercial entity (Maps of Cancer Mortality Rates in Spain), published by the media organisation El Pais^[https://elpais.com/].

### Geographical coverage

Identified cancer atlases covered geographies from all around the world, four were global atlases. Most focussed on single nations, the United States was considered by eleven atlases. The United Kingdom was considered by seven, followed by three of Australia, two of Canada, and one of each from Switzerland, Germany, Norway. One atlas covered the European Union.
Not all maps had a national focus and ten covered a region or state rather than an entire nation. The states or counties/regions covered were South Australia (AUS), Queensland (AUS), Ontario (CAN), Valencia (Spain), Pennsylvania county Massachusetts (US), New Hampshire (US), Cape Cod (US), Missouri (US), Florida (US), New York State (US) and Arizona (US).

### Statistical uncertainty in public facing atlases
